---
title: "PDA Project1"
author: 'Due: October 8 by 11.59pm'
date: "`r Sys.Date()`"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, 
                      warning=FALSE, 
                      error=FALSE, 
                      echo = FALSE, 
                      fig.pos = "H" ,
                      fig.align = 'center')
```


```{r load required libraries}

# libraries to be used 
#install.packages("kableExtra")
library(gtsummary)
library(tidyverse)
library(kableExtra)
library(knitr)
library(tidyr)
library(naniar)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tableone)
library(dplyr)
library(lattice)
library(reshape2)
library(formatR)
#library(summarytools)
#library(broom)

```


```{r}

# Data cleaning

# variable transformation
project_data$psex <- as.factor(project_data$psex)
project_data$plang <- as.factor(project_data$plang)
project_data$pethnic <- as.factor(project_data$pethnic)
project_data$paian <- as.factor(project_data$paian)
project_data$pasian <- as.factor(project_data$pasian)
project_data$pnhpi <- as.factor(project_data$pnhpi)
project_data$pblack <- as.factor(project_data$pblack)
project_data$pwhite <- as.factor(project_data$pwhite)
project_data$prace_other <- as.factor(project_data$prace_other)
project_data$employ <- as.factor(project_data$employ)
project_data$pedu <- as.factor(project_data$pedu)
project_data$childasd <- as.factor(project_data$childasd)
project_data$nidaalc <- as.factor(project_data$nidaalc)
project_data$nidatob <- as.factor(project_data$nidatob)
project_data$nidapres  <- as.factor(project_data$nidapres)
project_data$nidaill <- as.factor(project_data$nidaill)
project_data$mom_smoke_16wk <- as.factor(project_data$mom_smoke_16wk)
project_data$mom_smoke_22wk <- as.factor(project_data$mom_smoke_22wk)
project_data$mom_smoke_32wk <- as.factor(project_data$mom_smoke_32wk)
project_data$mom_smoke_pp1 <- as.factor(project_data$mom_smoke_pp1)
project_data$mom_smoke_pp2 <- as.factor(project_data$mom_smoke_pp2)
project_data$mom_smoke_pp12wk <- as.factor(project_data$mom_smoke_pp12wk)
project_data$mom_smoke_pp6mo <- as.factor(project_data$mom_smoke_pp6mo)
project_data$smoke_exposure_6mo <- as.factor(project_data$smoke_exposure_6mo)
project_data$smoke_exposure_12mo <- as.factor(project_data$smoke_exposure_12mo)
project_data$smoke_exposure_2yr <- as.factor(project_data$smoke_exposure_2yr)
project_data$smoke_exposure_3yr <- as.factor(project_data$smoke_exposure_3yr)
project_data$smoke_exposure_4yr <- as.factor(project_data$smoke_exposure_4yr)
project_data$tsex <- as.factor(project_data$tsex)
project_data$language <- as.factor(project_data$language)
project_data$tethnic <- as.factor(project_data$tethnic)
project_data$alc_ever <- as.factor(project_data$alc_ever)
project_data$mj_ever <- as.factor(project_data$mj_ever)
project_data$e_cig_ever <- as.factor(project_data$e_cig_ever)
project_data$cig_ever <- as.factor(project_data$cig_ever)
project_data$taian <- as.factor(project_data$taian)
project_data$tasian <- as.factor(project_data$tasian)
project_data$tnhpi <- as.factor(project_data$tnhpi)
project_data$tblack <- as.factor(project_data$tblack)
project_data$twhite <- as.factor(project_data$twhite)
project_data$trace_other <- as.factor(project_data$trace_other)
project_data$smoke_exposure_5yr <- as.factor(project_data$smoke_exposure_5yr)
project_data$momcig <- as.numeric(project_data$momcig)
project_data$mom_numcig <- as.numeric(project_data$mom_numcig)
project_data$income <- as.numeric(project_data$income)
project_data$cotimean_34wk <- as.numeric(project_data$cotimean_34wk)
project_data$cotimean_pp6mo <- as.numeric(project_data$cotimean_pp6mo)
project_data$cotimean_pp6mo_baby <- as.numeric(project_data$cotimean_pp6mo_baby)

#project_data$mom_smoke_16wk <- as.numeric(project_data$mom_smoke_16wk)
#project_data$cmom_smoke_22wk <- as.numeric(project_data$mom_smoke_22wk)
#project_data$mom_smoke_32wk <- as.numeric(project_data$mom_smoke_32wk)


# blank values to NA
project_data[project_data == ""] <- NA

# change the income input 
project_data$income[6] <- 250000
project_data$income[1] <- 76000

# handle the outlier value on momcig 
project_data$momcig[31] <- 4

# replace the erronously recorded value with NA
project_data[1, "mom_numcig"] <- 2
project_data[26, "mom_numcig"] <- NA # 44489
project_data[37, "mom_numcig"] <- 23  # 20-25
project_data[47, "mom_numcig"] <- NA  # none

# update 'num_cig_30' column to show true values
project_data <- project_data %>% mutate(num_cigs_30 = 
                                          case_when(cig_ever == 0 ~ 0,                               
                                          cig_ever == 1 ~ num_cigs_30,
                                          TRUE ~ NA))
# update 'num_e_cig_30' column
project_data <- project_data %>% mutate(num_e_cigs_30 = case_when(e_cig_ever == 0 ~ 0,
                                        e_cig_ever == 1 ~ num_e_cigs_30,
                                        TRUE ~ NA))
# update 'num_mj_30' column
project_data <- project_data %>% mutate(num_mj_30 = case_when(mj_ever == 0 ~ 0,
                                        mj_ever == 1 ~ num_mj_30,
                                        TRUE ~ NA))
# update 'num_alc_30' column
project_data <- project_data %>% mutate(num_alc_30 = case_when(alc_ever == 0 ~ 0,
                                        alc_ever == 1 ~ num_alc_30,
                                        TRUE ~ NA))

```

###   **2.2 Missing Data**


```{r}
# summary statistics for missing data 
missing_data <- data.frame(
  Variable = names(project_data),
  Missing_Count = sapply(project_data, function(x) sum(is.na(x)))
)
missing_data$Percentage_Missing <- missing_data$Missing_Count / nrow(project_data) * 100
missing_data <- missing_data %>%
  arrange(desc(Percentage_Missing))

# Remove row names
rownames(missing_data) <- NULL

# variables with missing values above 25%
missing_data_only <- missing_data %>%
  filter(Percentage_Missing > 25)
project_missing <- naniar::miss_var_summary(project_data, order = FALSE) 
project_missing$pct_miss <- round(project_missing$pct_miss, 2)
missing_data_only$Missing_Count <- round(missing_data_only$Missing_Count, 2)

missing_data_only$Percentage_Missing <- round(missing_data_only$Percentage_Missing, 2)

missing_data_only %>% 
kable(caption = "Variables with above 25% Missing Data", 
      col.names = c("Variable", "Proportion (%)", "n"),
        digits = 4,
        booktabs = TRUE) %>% 
 kable_styling(full_width = F,latex_options = c("HOLD_position", "striped"),
              font_size=9)

```



##    **3. Analysis**

###   **3.1 Demographic summary statistics**

```{r}

# summary table for demographic characteristics
tbl_summary_demographic <-  project_data %>%
  select(page,pedu,employ,tage , tsex, prace_other, trace_other) %>%
  tbl_summary(missing = "no", 
              label = list(page = "Mom Age", pedu = "Mom Education", employ = "Mom Employed",tage = "Child Age" , tsex = "Child Sex", prace_other = "Mom Other Race", trace_other = "Child Other Race")) 

# change tbl_summary to a data frame
data_frame <- tbl_summary_demographic %>%
  as_tibble()
column_names <- c("Variable", "N = 49") 
colnames(data_frame) <- column_names

# for nicer display
kable_tbl <- data_frame %>%
 kable(caption = "Demographic Characteristics", 
      col.names = c("Variable", "N = 49"),
        digits = 4,
        booktabs = TRUE) %>% 
 kable_styling(full_width = F,latex_options = c("HOLD_position", "striped"),
              font_size=9)

kable_tbl

```


```{r Interrelatedness of SDP and ETS}
# Standardize cotinine variable for mom and child
mom_cotimean_34_z <- abs(log(scale(project_data$cotimean_34wk)))
mom_cotimean_6mo_z <- abs(log(scale(project_data$cotimean_pp6mo)))
child_cotimean_6mo_z <-abs(log(scale(project_data$cotimean_pp6mo_baby)))

# calculate composite scores for SDP variable
project_data$SDP <- with(project_data, 
                         rowMeans(cbind(mom_numcig, mom_smoke_16wk, 
                                        mom_smoke_22wk, mom_smoke_32wk, 
                                        mom_cotimean_34_z),na.rm = TRUE))

# calculate composite scores for ETS variable
project_data$ETS <- with(project_data, 
                         rowMeans(cbind(mom_smoke_pp1, mom_smoke_pp2, 
                                        mom_smoke_pp12wk, mom_smoke_pp6mo, 
                                        smoke_exposure_6mo, smoke_exposure_12mo, 
                                        smoke_exposure_2yr, smoke_exposure_3yr,
                                        smoke_exposure_4yr,
                                        smoke_exposure_5yr,mom_cotimean_6mo_z,
                                        child_cotimean_6mo_z),na.rm = TRUE))


```

```{r, include=FALSE}
# correlation matrix for SDP  and ETS
correlation_matrix_b <- cor(project_data[c("SDP", "ETS")], use ="complete.obs")

# reshape the correlation matrix for ggplot
melted_cor_matrix_b <- melt(correlation_matrix_b)

# correlation matrix visual
ggplot(melted_cor_matrix_b, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                        name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 1,
                                   size = 5, hjust = 1), 
        axis.text.y = element_text(size = 5)) +
  labs(x = NULL, y = NULL, fill = "Correlation") +
  coord_fixed() +
  ggtitle("SDP and ETS matrix")+
  labs(caption = "Figure 1: SDP vs ETS Correlation Matrix") 
```


```{r}
# SDP versus ETS
ggplot(project_data, aes(x = SDP, y = ETS)) +
  geom_point(size = 1, alpha = 0.7, color = "green") +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, 
              linetype = "solid", color = "red") +
  labs(title = "Effect of Environmental tobacco smoking on SR",
       x = "Smoking During Pregnancy",
       y = "Environmental Tobacco Smoking") +
  theme_minimal() +
  labs(caption = "Figure 1: Association between ETS and SDP")
```


```{r}
# correlation matrix for SR/EXT
correlation_matrix_out <- cor(project_data[c("bpm_att", "bpm_ext", "bpm_att_p", "bpm_ext_p", "swan_inattentive", "swan_hyperactive","erq_exp","erq_cog","bpm_int", "bpm_int_p")], use ="complete.obs")

# reshape the correlation matrix for ggplot
melted_cor_matrix_out <- melt(correlation_matrix_out)

# correlation matrix visual
ggplot(melted_cor_matrix_out, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       #breaks = seq(-1, 1, by = 0.5)) +
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 1,
                                   size = 5, hjust = 1), 
        axis.text.y = element_text(size = 5)) +
  labs(x = NULL, y = NULL, fill = "Correlation") +
  coord_fixed() +
  ggtitle("Matrix for Outcome ")+
  labs(caption = "Figure 2: Correlation Matrix") 
```


```{r}

# Standardize SR, EXT variables for child
bpm_att_z = abs(scale(project_data$bpm_att))
bpm_ext_z = abs(scale(project_data$bpm_ext))
bpm_int_z = abs(scale(project_data$bpm_int))
erq_cog_z = abs(scale(project_data$erq_cog))
erq_exp_z = abs(scale(project_data$erq_exp)) 
erq_cog_a_z = abs(scale(project_data$erq_cog_a)) # exclude parent
erq_exp_a_z = abs(scale(project_data$erq_exp_a)) # exclude parent
bpm_att_p_z = abs(scale(project_data$bpm_att_p))
bpm_ext_p_z = abs(scale(project_data$bpm_ext_p))
bpm_int_p_z = abs(scale(project_data$bpm_int_p))
bpm_att_a_z = abs(scale(project_data$bpm_att_a)) # exclude parent
bpm_ext_a_z = abs(scale(project_data$bpm_ext_a)) # exclude parent
bpm_int_a_z = abs(scale(project_data$bpm_int_a)) # exclude parent
swan_hyperactive_z = abs(scale(project_data$swan_hyperactive))
swan_inattentive_z = abs(scale(project_data$swan_inattentive))

# calculate composite scores for self regulation variable
project_data$SR <- with(project_data, 
                        rowMeans(cbind(erq_cog_z, erq_exp_z, bpm_int_p_z, 
                                       bpm_int_z),na.rm = TRUE))

# calculate composite scores for externalizing variable
project_data$EXT <- with(project_data, 
                         rowMeans(cbind(bpm_att_z, bpm_ext_z, bpm_att_p_z,
                                        bpm_ext_p_z, swan_inattentive_z, swan_hyperactive_z),na.rm = TRUE))

# composite for substance use variable
project_data$SU <- with(project_data,
                        ifelse(cig_ever == 1 | e_cig_ever == 1 | 
                                 mj_ever == 1 | alc_ever == 1, 1, 0))


## variables for the outcome variables
df_outcomes = project_data %>% 
  select(bpm_att, bpm_ext, bpm_att_p, bpm_ext_p, swan_inattentive,erq_cog, erq_exp, swan_hyperactive)

# correlation matrix for interrelatedness externalizing variables
correlation_matrix <- cor(project_data[c("bpm_att", "bpm_ext", 
                                         "bpm_att_p","bpm_ext_p", "swan_inattentive", 
                                         "swan_hyperactive")], use = "pairwise.complete.obs")

# correlation matrix for interrelatedness self regulation
correlation_matrix2 <- cor(project_data[c("erq_exp", "erq_cog", "bpm_int", "bpm_int_p")], 
                           use = "pairwise.complete.obs")


```

###   3.3.2 Interrelatedness of the externalizing and self regulate behaviors in children
```{r, include=FALSE}
# correlation matrix for SR/EXT
#correlation_matrix <- cor(project_data[c("EXT", "SU", "SR", 
                            # "SDP", "EXT")], use ="complete.obs")
# correlation matrix for SR/EXT
correlation_matrix_c <- cor(project_data[c("EXT", "SR")], use ="complete.obs")

# reshape the correlation matrix for ggplot
melted_cor_matrix_c <- melt(correlation_matrix_c)

# correlation matrix visual
ggplot(melted_cor_matrix_c, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       #breaks = seq(-1, 1, by = 0.5)) +
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 1,
                                   size = 5, hjust = 1), 
        axis.text.y = element_text(size = 5)) +
  labs(x = NULL, y = NULL, fill = "Correlation") +
  coord_fixed() +
  ggtitle("EXT vs SR Correlation Matrix")+
  labs(caption = "Figure 3: EXT/SR Matrix") 
```

```{r}
# ETS on self regulation
ggplot(project_data, aes(x = ETS, y = SR)) +
  geom_point(size = 1, alpha = 0.7, color = "green") +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, 
              linetype = "solid", color = "red") +
  labs(title = "Effect of Environmental tobacco smoking on SR",
       x = "Environmental Tobacco Smoking",
       y = "Self Regulation Problems") +
  theme_minimal() +
  labs(caption = "Figure 3: Effect of Environmental Tobacco Smoking on self Regulation")
```


```{r Effect of ETS on Externalizing problems}

# ETS on Externalizing problems
ggplot(project_data, aes(x = ETS, y = EXT)) +
  geom_point(size = 1, alpha = 0.7, color = "green") +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, 
              linetype = "solid", color = "red") +
  labs(title = "Effect of Environmental tobacco smoking on EXT",
       x = "Environmental Tobacco Smoking",
       y = "Externalizing problems") +
  theme_minimal() +
  labs(caption = "Figure 4: Effect of Environmental Tobacco Smoking on Externalizing problems")

```


```{r Fitting Regression Models}
# model selection for externalizing problem as the outcome
# fit first model with all covariates 
model3 = lm(EXT ~ SDP +  ETS + page + pethnic + paian + pnhpi + pwhite + prace_other + employ + 
              pedu + income + tage + tethnic + taian + tblack + twhite + trace_other, data = project_data)


# drop 'tethnic' 
model3.1 = lm(EXT ~ SDP +  ETS + page + pethnic + paian + pnhpi + pwhite + prace_other + employ + pedu + income + tage + taian + tblack + twhite + trace_other, data = project_data)


# drop 'trace_other'
model3.2 = lm(EXT ~ SDP +  ETS + page + pethnic + paian + pnhpi + pwhite + prace_other + employ +
                pedu + income + tage  + taian + tblack + twhite, data = project_data)


# drop 'pwhite'
model3.3 = lm(EXT ~ SDP +  ETS + page + pethnic + paian + pnhpi + twhite + prace_other + employ + pedu + income + tage  + taian + tblack , data = project_data)

# drop 'taian'
model3.4 = lm(EXT ~ SDP +  ETS + page + pethnic + paian + pnhpi + pwhite + prace_other + employ + pedu + income + tage  + tblack , data = project_data)

# drop 'pnhpi'
model3.5 = lm(EXT ~ SDP +  ETS + page + pethnic + paian + pedu + pwhite + prace_other + employ + income + tage  + tblack , data = project_data)


# drop 'income'
model3.6 = lm(EXT ~ SDP +  ETS + page+ pethnic+ paian + pedu + pwhite + prace_other + employ +
                tage  + tblack , data = project_data)


# final model # adjust for page
model3.6 = lm(EXT ~ SDP +  ETS + pethnic+ paian + pedu + pwhite + prace_other + employ +
                tage  + tblack , data = project_data)
#summary(model3.6)

logm1 = glm(as.factor(SU) ~ SDP +  ETS + page + pethnic + paian + pnhpi + pwhite + prace_other + employ + pedu + income + tage + tethnic + taian + tblack + twhite + trace_other,family = binomial,  data = project_data)
#summary(logm)

logm2 = glm(as.factor(SU) ~ SDP +  ETS  + prace_other +tblack   ,family = binomial,  data = project_data)
#summary(logm2)
```


\newpage
##    References

1. McDaniel JT, Lockhart E, Albright DL, et al. Cigarette Use among Parents with Children in the Home: A Comparison of Civilian Parents and Parents with a Military Connection. Environmental Health Insights. 2022;16:1-61 https://doi.org/10.1177/11786302221137220

\newpage
# Code Appendix: 

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```

